---
title: "Introduction to traveltimeHMM: making reliable travel time predictions on road networks"
author: "Mohamad Elmasri, Aurélie Labbe, Denis Larocque, Laurent Charlin and Éric Germain"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to traveltimeHMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```
# Introduction
In the new world of data, decision-making is becoming more data-driven then ever. This is increasingly true for urban transportation, where the ubiquitous expected time of arrival (ETA) is heavily used. Providing reliability measure for ETA's is likely to improve our understanding of travel time and to provide systems with better decision-making metrics. This `traveltimeHMM` package provides various methods to estimate the distribution of travel time on a given route. This is done by by leveraging GPS data from moving vehicles on road segments, with flexible temporal binning, to provide travel time estimate for complete routes. 

Impelemented methods include the proposed TRIP model (and its variates) in Woodard et al. (2017). Other methods methods are in the pipeline. 

This is a work in progress. If you found a bug, please report it [here](https://github.com/melmasri/traveltimeHMM/issues), if would like to suggest (or provide) a feature please refer to the [main repository](https://github.com/melmasri/traveltimeHMM).


# Example data set: `tripset`

This package includes a small data set (`tripset`) that aggregates map-matched anonymized mobile phone GPS data collected in Quebec city in 2014 using the Mon Trajet smartphone application developed by [Brisk Synergies Inc](https://brisksynergies.com/). The precise duration of the time period is kept confidential.

We use `tripset` to explore the functionalities offered in this package. Data similar to `tripset` can be generated using other sources of GPS data, however, a preprocessing stage is necessary to map-match and aggregate the information up to the road segment (link) level. Such preprocessing is already performed on `tripset`.

We stress that `traveltimeHMM` does not perform this preprocessing stage, and that the software cannot make use of raw GPS data.


```{r}
library(traveltimeHMM)
data(tripset)
head(tripset)
```

Travel data is organized around the notions of *trips* and *links*.  Links are road segments each with well-defined beginning and end points and which can be traversed. A vehicle performs a trip when it travels from a start point to an end point through a sequence of links.  Thus trips can be considered as ordered sequences of links. `tripset` includes data for a collection of trips.

- Field `tripID` contains each *trip*'s ID, whereas field `linkID` contains the IDs of each *link* making up a trip.  Both fields need to be numerical.  It is assumed that, in the data set, all trips are grouped together and all links of a given trip appear in the order in which they are traversed^[No verification is performed to that effect.].  A given link corresponds to some physical entity such as a road segment, a portion of a road segment, or any combination of those.  Hence it is expected that links are used in more than one trip.

- Field `timeBin` (character string) refers to the time "category" when the traversal of a given link occurred.  Time bins should reflect as much as possible time periods of the week encompassing similar traffic classes.  In `tripset` we find five time bins: `Weekday`, `MorningRush`, `EveningRush`, `EveningNight` and `Weekendday`.  Field `time` served as a basis for determining the time bin for each link traversal, as illustrated in table 1.  *### TO DO: include a mention regarding the need to provide the conversion algorithm from time stamps to time bins.  See issue #17.  ÉG 2019/07/08 ###* 

```{r echo = FALSE, results='asis'}
library(knitr)
library(kableExtra)
y <- c("Mon - Fri outside rush hour", "Mon - Fri, 7AM - 9AM", "Mon - Fri, 3PM - 6PM", "Sat 9AM - 9PM + Sun 9AM - 7PM", "Sat - Sun otherwise")
x <- c("Weekday", "MorningRush", "EveningRush", "Weekendday", "EveningNight")
z <- data.frame(cbind(y, x))
names(z) <- c("Period of the week", "Time bin")
knitr::kable(z, caption = "Table 1 - Example from `tripset` : Time bin by period of the week", escape = FALSE)
```

- Field `logspeed` contains the natural logarithm of the *speed* (in meter/second) of traversal of each link.  This information is central to the estimation algorithm.

- Field `traveltime` refers to the traversal time (in seconds) of each link.  This field is mostly for reference and is not used directly in the package.

- Field `length` refers to the length (in meters) of each link.  This information is used by the prediction algorithm.

- Field `time` gives the start time in POSIXct format for the traversal of a given link.  Individual datums are not used directly; however, the start time of the very first traversal is likely to be useful for providing the start time of a whole trip to the prediction function.

# Statistical models

At the core of `traveltimeHMM` is a statistical model that aims at determining the probability distribution of the logarithm of *speeds* for each combination of *link* and *time bin*, on the basis of data just like that we just described.  We will first briefly present the model and its parameters.  We refer to Woodard et al. (2016) for an extensive description of the model.  We will subsequently present the four specific model types that are currently implemented, each model utilizes a different set of parameters.

## Model parameters

The most general model is the `trip-HMM`, which estimates the mean and standard deviation of the *speed probability distribution* for each combination of road link, time bin, and congestion state.  This latter concept involved defining a finite number of *hidden states* that reflect traffic fluidity of a trip at a given place and time, for instance the fact that one is driving downtown during rush hour (or not), they might (or might not) fall into congestion at a specific link and time. Hence, conngestion level as a property of the individual trip, rather than as a property of the link and time. In our example, we chose to define two congestions states which we could implicitly call "not-congested" and "congested".

In order to deal properly with the fact that the congestion states are unobserved, the model needs to also estimate parameters that define the *probability* of a combination of road link and time bin *reaching a given state*.  These estimates are computed using a Markov model.  Finally, the model estimates a *trip effect* that captures variability in speed due to trip-specific conditions (e.g. driver habits).

The estimate parameters of model `trip-HMM` are:

* for each combination of road link $j$, time bin $b$ and congestion state $q$, we get:

    * The mean $\mu_{j,b,q}$ and standard deviation $\sigma_{j,b,q}$ of the log-speed probability distribution (assuming a Gaussian distribution);
    
    * A vector $\gamma_{j,b}(q)$ which shows the probability of the congestion state $q$ for trips starting on link $j$ during $b$; and

    * A transition matrix $\Gamma_{j,b}(\tilde{q},q)$ for the congestion state $q$ on $j$ and $b$, conditional on $\tilde{q}$, the congestion state in the previous link.
    
* We also get the trip effect $\log(E_i) \sim N(0,\tau^2)$ for each trip $i$, as well as its standard deviation $\tau$.

The other three model types are special cases of the most general model.  The `HMM` model, which is the default, assumes that the trip effect is fixed to 1 ($E_i=1$). The `trip` model
considers the existence of a single traffic state (does not use the hidden states abstraction); under this model the parameters $\gamma$ and $\Gamma$ remain undefined.  Finally, the `no-dependence` model is the most basic of the four, combines the restrictions of the `HMM` and `trip` model types and only takes into account parameters $\mu_{j,b,q}$ and $\sigma_{j,b,q}$ for the speed probability distribution, hence not trip effect and no hidden Markov states. Table 2 provides a summary of the characteristics of each model type.

```{r echo = FALSE, results='asis'}
library(knitr)
library(kableExtra)
x1 <- c("trip-HMM", "HMM (default)", "trip", "no-dependence")
x2 <- c("YES", "YES", "NO", "NO")
x3 <- c("YES", "NO", "YES", "NO")
z <- data.frame(cbind(x1, x2, x3))
names(z) <- c("Model type", "Hidden Markov model (`HMM`)", "Trip effect model (`trip`)")
knitr::kable(z, caption = "Table 2 - Model typology", escape = FALSE, align = 'lcc') %>%
  kable_styling(full_width = F) %>%
  column_spec(1, width = "2cm") %>%
  column_spec(2:3, width = "3cm")
```

## Model estimation

Model estimation is performed using Algorithm 1 from Woodard et al. (2016)  The estimation algorithm takes incoming data on speeds (on a logarithmic scale), trips, road links and time periods, and calculates a set of parameter estimates for the model.

Calculations are performed iteratively as is typical of non-linear optimization algorithms.  The estimates for each parameter are updated at each iteration, and the algorithm stops when either the variation in parameter estimates between two successive iterations is below a predefined threshold value, or a predefined maximum number of iterations is reached.

## Calling `traveltimeHMM`

The estimation algorithm can be executed by calling the `traveltimeHMM` function.  The function's interface is as follows.

```{r eval=FALSE}
traveltimeHMM(logspeeds = NULL,
              trips = NULL,
              timeBins = NULL,
              linkIds = NULL,
              tripframe = NULL,
              nQ = 1L,
              model = c("HMM", "trip-HMM", "trip", "no-dependence"),
              tol.err = 10,
              L = 10L,
              max.it = 20,
              verbose = FALSE,
              max.speed = NULL,
              seed = NULL,
              tmat.p = NULL,
              init.p = NULL,
              debug = FALSE)
```

Function `traveltimeHMM` can be called either using *explicit parameters* or using a *composite object* incorporating the main parameters, as follows.  As an example of the first calling approach, we extract columns ```logspeeds```, ```trips```, ```timeBins``` and ```linkIds``` from the dataset and pass them explicitly:
```{r eval=FALSE}
# Example: calling traveltimeHMM with explicit parameters
fit <- traveltimeHMM(logspeeds = tripset$logspeed,
                     trips = tripset$tripID,
                     timeBins = tripset$timeBin,
                     linkIds = tripset$linkID,
                     nQ = 2,
                     max.it = 20,
                     model = "HMM" # this is the default value)
```

As an example of the second calling approach, we construct and pass a single data frame object that contains those same four columns:
```{r eval=FALSE}
fit <- traveltimeHMM(data = tripset,
                     nQ = 2,
                     max.it = 20,
                     model = "HMM")
```

Table 3 provides a description for each parameter.

```{r echo = FALSE, results='asis'}
library(knitr)
library(kableExtra)

r <-       c("`logspeeds`", "A numeric vector of speed observations (in km/h) on the (natural) log-scale.  Needs to be provided if and only if `tripframe` is NULL.  Default is NULL.")

r <- rbind(r,
           c("`trips`", "An integer or character vector of trip ids for each observation of `speed`.  Needs to be provided if and only if `tripframe` is NULL.  Default is NULL."))

r <- rbind(r,
           c("`timeBins`", "A character vector of time bins for each observation of `speed`.  Needs to be provided if and only if `tripframe` is NULL.  Default is NULL."))

r <- rbind(r,
           c("`linkIds`", "A vector of link ids (route or way) for each observation of `speed`.  Needs to be provided if and only if `tripframe` is NULL.  Default is NULL."))

r <- rbind(r,
           c("`tripframe`", "A data frame or equivalent object that contains one column for each of test.  Default is NULL.  Mutually exclusive with the full joint set of `logspeeds`, `trips`, `timeBins` and `linkIds`."))

r <- rbind(r,
           c("`nQ`", "An integer corresponding to the number of different congestion states that the traversal of a given link can take corresponding to $|\\{1, ..., Q\\}|$.    Models of the HMM family require `nQ` $\\geq 2$ whilst other models require exactly `nQ` $= 1$.  Default is $1$."))

r <- rbind(r,
           c("`model`", "Type of model as string.  Can take one of '`trip-HMM`', '`HMM`', '`trip`' or '`no-dependence`'.  Default is '`HMM`'.  See Table 3 for details."))

r <- rbind(r,
           c("`tol.err`", "A numeric variable representing the threshold under which the estimation algorithm will consider it has reached acceptable estimate values.  Default is $10$."))

r <- rbind(r,
           c("`L`", "An integer minimum number of observations per factor (`linkIds` x `timeBins`) to estimate the parameter for.  Default is $10$.  See section on imputation below."))

r <- rbind(r,
           c("`max.it`", "An integer for the upper limit of the iterations to be performed by the estimation algorithm.  Default is $20$."))

r <- rbind(r,
           c("`verbose`", "A boolean that triggers verbose output.  Default is `FALSE`."))

r <- rbind(r,
           c("`max.speed`", "An optional float for the maximum speed in km/h, on the linear scale (not the log-scale, unlike for `logspeeds`).  Default is `NULL` which in practice results in a maximum speed of 130 km/h.  # Change to default is 130.  ÉG 2019/07/22"))

r <- rbind(r,
           c("`seed`", "An optional float for the seed used for the random generation of the first Markov transition matrix and initial state vector.  Default is `NULL`. If not provided, then those objects are generated deterministically. The effect of `seed` is cancelled by tmat.p or init.p when provided."))

r <- rbind(r,
           c("`tmat.p`", "An optional Markov transition matrix $\\Gamma$ of size `nQ` \\times `nQ` with rows summing to $1$.  Default is `NULL`.  See section on Markov transition matrix and initial state vector below."))

r <- rbind(r,
           c("`init.p`", "An optional Markov initial state vector $\\gamma$ of size `nQ` with elements summing to $1$.  Default is `NULL`.    See section on Markov transition matrix and initial state vector below."))

r <- rbind(r,
           c("`debug`", "A boolean with value `TRUE` if we want debug information to be generated.  Default is `FALSE`."))

z <- as.data.frame(r[,2], row.names = r[,1])
names(z) <- ""
knitr::kable(z, caption = "Table 3 - Parameters for `traveltimeHMM`", escape = FALSE)
```

## Markov transition matrix and initial state vector below

If provided and valid, parameters `tmat.p` and `init.p` will be used as initial values for $\Gamma$ and $\gamma$.  Otherwise, the algorithm will assume initial equiprobability.  In our example, default initial values for the Markov transition matrix and initial state vector are:

<center>$\Gamma_{j,b}^{(0)}(\tilde{q},q) = \begin{bmatrix}
0.5 & 0.5 \\
0.5 & 0.5
\end{bmatrix}$ and $\gamma_{j,b}^{(0)}(q) = \begin{bmatrix}
0.5 \\
0.5
\end{bmatrix}$</center>

Each row and column of $\Gamma$ and each row of $\gamma$ correspond to different congestion states $q_1,q_2,....$  Given $j$ and $b$, $\Gamma_{row,col|j,b}$ represents the probability of reaching state $q_{col}$ from state $q_{row}$, whilst $\gamma_{row|j,b}$ represents the probability of beginning a trip with state $q_{row}$.

## A word on imputation

The implementation allows imputing for combinations of road link and time bin for which the lack of sufficient observations prevents reliable parameter estimation.  When imputation is required for a given combination, estimates for $\mu$, $\sigma$, $\Gamma$ and $\gamma$ are calculated on the basis of data available for the time bin involved *for all road links*.  This approach differs from the one in Woodard et al., where imputation is performed on the basis of road classification data (e.g. "arterial" or "primary collector road").  This implementation does not handle road classification data.

Imputation is performed in the following three cases:

- Case 1: when a combination has fewer than $L$ *total observations*;

- Case 2: when a combination has fewer than $L$ *initial state observations only*;

- Case 3: when a combination has only initial states.

The number of observations specified in parameter `L` determines the threshold below which imputation occurs for cases 1 and 2.

## Output from execution

Below is some typical output from executing `travetimeHMM`.  One can find:

* the maximum speed on the road network;
* the number of trips, road links and time bins;
* the number of iterations to be executed;
* an estimate of the total execution time, based on the time it takes for executing a firts iteration.

Error messages are described in the appendix. # (To be done later... ÉG 2019/07/25)

```{r}
#> max.speed is not specified, setting at default value: 130 km/h
#> Model HMM with 4914 trips over 13235 roads and 5 time bins...
#> Expected completion of 20 iterations in 105 secs
#> Reached maximum number of iterations
```

#### Return values

The execution of `traveltimeHMM` returns a list of the parameters in Table 4. 

```{r echo = FALSE, results='asis'}
library(knitr)
library(kableExtra)

r <-       c("`factors`", "A factor of interactions (`linkIds` x `timeBins`) of length `nObs`.  Factors are in the format 'linkId.timeBin'.")

r <- rbind(r,
           c("`trip`", "A factor of trip IDs."))

r <- rbind(r,
           c("`tmat`", "A transition matrix with rows corresponding to `levels(factors)`, and with columns being the row-wise transition matrix of that factor. For example, `matrix(tmat[1,], ncol = nQ, nrow = nQ, byrow = TRUE)` is the transition matrix $\\hat{\\Gamma}$ of `levels(factors)[1]`.  NULL if hidden Markov modelling is not handled by the selected model type."))

r <- rbind(r,
           c("`init`", "An initial state probability matrix with rows corresponding to `levels(factors)`, and columns to the `nQ` states.  For example, `init[1,]` gives $\\hat{\\gamma}^\\top$ for `levels(factors)[1]`.  NULL if hidden Markov modelling is not handled by the selected model type."))

r <- rbind(r,
           c("`sd`", "A matrix of standard deviations estimates for the  natural logarithm of the speed (in km/h), with rows corresponding to `levels(factors)`, and columns to standard deviation estimates $\\hat{\\sigma}$ for the `nQ` states."))

r <- rbind(r,
           c("`mean`", "A matrix of mean estimates for the  natural logarithm of the speed (in km/h), with rows corresponding to `levels(factors)`, and columns to mean estimates $\\hat{\\sigma}$ for the `nQ` states.  Speed"))

r <- rbind(r,
           c("`tau`", "A numeric variable for the standard deviation estimate $\\hat{\\tau}$ for the trip effect $\\log(E)$.  Equals $1$ if trip effect is not handled by the selected model type."))

r <- rbind(r,
           c("`logE`", "A numeric vector of trip effect estimates $\\log(\\hat{E})$ corresponding to `levels(trip)`.  Values are set to $0$ if trip effect is not handled by the selected model type.  Units are the same as for `logspeeds`."))

r <- rbind(r,
           c("`nQ`", "An integer corresponding to the number of different congestion states, equal to the parameter `nQ` that was passed in the function call."))

r <- rbind(r,
           c("`nB`", "An integer corresponding to the number of unique time bins."))

r <- rbind(r,
           c("`nObs`", "An integer corresponding to the number of observations."))

r <- rbind(r,
           c("`model`", "Type of model as string.  Same as parameter `model` that was passed in the function call."))
z <- as.data.frame(r[,2], row.names = r[,1])
names(z) <- ""
knitr::kable(z, caption = "Table 4 - Components of the list object returned by `traveltimeHMM`", escape = FALSE)
```

## Prediction

Prediction is performed using Algorithm 2 from Woodard et al.  Each prediction is obtained by simulation on the basis of the parameter estimates supplied to the algorithm.  In short, the algorithm starts with the first road link and determines by simulating the initial congestion state considering the time bin and vector $\hat{\gamma}$.  The algorithm then samples the speed on the link given the congestion state and time bin as well as parameter estimates $\hat{\mu}$ and $\hat{\sigma}$.  Time after traversal of the link is then computed deterministically.  The process is repeated iteratively on the remaining links, this time using the information provided by $\hat{\Gamma}$, ultimately giving a point prediction for the travel time (in seconds) for the whole trip.

#### Calling `predict.traveltime`

The prediction algorithm is executed by calling the `predict.traveltime` function, which has the following interface:

```{r eval=FALSE}
predict(modelObject,
        tripdata,
        starttime = Sys.time(),
        nbRuns = 1000,
        logE = NULL # maybe logE = 0 ?  ÉG 2019/07/24)
```

Our example involves predicting the travel time for trip ID 2700 from the original data set, using the model parameters we just estimated, as follows.  We need to manually supply the start time by taking the very first link's time stamp.  Choosing `nbRuns = 1000` generates as many independent predictions of the total travel speed on the trip

```{r eval=FALSE}
# Extracting link traversal data for trip ID 2700.  Link traversal order is preserved.
single_trip <- subset(tripset, tripID==2700)

pred <- predict.traveltime(modelObject = fit,
                           tripdata = single_trip,
                           starttime = single_trip$time[1],
                           nbRuns = 1000)
```
                           

Table 5 provides a description for each parameter.

```{r echo = FALSE, results='asis'}
library(knitr)
library(kableExtra)

r <-       c("`modelObject`", "A list object corresponding to the return value of function `traveltimeHMM`.")

r <- rbind(r,
           c("`tripdata`", "A data frame of road links with information on each link's traversal. Columns minimally includes objects `linkID` and `length`, and the latter must have the same length. Rows must be in chronological order."))

r <- rbind(r,
           c("`starttime`", "The start date and time for the very first link of the trip, in `POSIXct` format. Default is the current date and time."))

r <- rbind(r,
           c("`nbRuns`", "Number of simulation runs. Default is $1000$."))

r <- rbind(r,
           c("`logE`", "A numeric representing the point estimate of the mean trip effect for the speed in km/h, on the log-scale.  (Hence, speeds and trip effects are to be added *on the logarithmic scale* as prescribed in Woodard et al.)  `logE` normally needs to be a vector of numerics of size `nbRuns`. If a single numeric value is supplied, it will be replicated into a vector. If `logE` is `NULL` the function will use either a vector of simulated values (if the model is from the trip family), or a vector of $0$ otherwise. Default is NULL. NOTE: when simulating values for the vector, the value for $\\tau$ is taken from the model object."))

z <- as.data.frame(r[,2], row.names = r[,1])
names(z) <- ""
knitr::kable(z, caption = "Table 5 - Parameters for `predict.traveltime`", escape = FALSE)
```

## Return values

`predict.traveltime` returns a vector of size nbRuns of numerics representing the point prediction of total travel time, in seconds, for each run.

```{r echo=FALSE}
load("predHMM.RData")
load("single_trip.RData")
```

```{r}
head(pred)
```

Figure 1 shows the travel time distribution thus obtained.  The observed travel time for the same trip in the data set is 887.7327 seconds, whilst the mean of the 1 000 predictions for our simulation was 861.2652 seconds.

```{r results = 'asis', fig.cap = "Figure 1 - Travel time distribution for trip ID 2700 for 1 000 runs using the test data set with an HMM model.", fig.width=4.5, fig.height=3}
hist(pred, freq =FALSE)
abline(v = mean(pred), lty=2, lwd=2)
abline(v=sum(single_trip$traveltime), lty=2, lwd=2, col='red')
```

# Model type comparison

Figure 2 shows the histograms of travel time for each of the four model types for the same example.  In this very case, the `HMM` model type seems to converge better than the other three to the observed travel time.

```{r echo = FALSE, results = 'asis', fig.cap = "Figure 2 - Comparison of travel time distribution for the four model types for the same example.", fig.width=6, fig.height=4}
par(mar=c(2,2,2,2))
par(mfrow=c(2,2))

load("predTRIP_HMM.RData")
hist(pred, freq =FALSE, main = "trip-HMM")
abline(v = mean(pred), lty=2, lwd=2)
abline(v=sum(single_trip$traveltime), lty=2, lwd=2, col='red')

load("predHMM.RData")
hist(pred, freq =FALSE, main = "HMM")
abline(v = mean(pred), lty=2, lwd=2)
abline(v=sum(single_trip$traveltime), lty=2, lwd=2, col='red')

load("predTRIP.RData")
hist(pred, freq =FALSE, main = "trip")
abline(v = mean(pred), lty=2, lwd=2)
abline(v=sum(single_trip$traveltime), lty=2, lwd=2, col='red')

load("predNoDependence.RData")
hist(pred, freq =FALSE, main = "no-dependence")
abline(v = mean(pred), lty=2, lwd=2)
abline(v=sum(single_trip$traveltime), lty=2, lwd=2, col='red')

```

# References

Woodard, D., Nogin, G., Koch, P., Racz, D., Goldszmidt, M., Horvitz, E., 2017.  "Predicting travel time reliability using mobile phone GPS data".  *Transportation Research Part C*, 75, 30-44.  <http://dx.doi.org/10.1016/j.trc.2016.10.011>


