---
title: "An introduction to traveltimeHMM : Making reliable travel time predictions for routes on a road networks"
author: "Mohamad Elmasri, Aurélie Labbe, Denis Larocque, Laurent Charlin and Éric Germain"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to traveltimeHMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Travel time prediction has many uses for individuals and organizations wishing to estimate the time it takes to complete a given route, but also to choose the best route available for a given trip.  Predictions are likely to be more useful if they include some reliability measure for the expected travel time.  The `traveltimeHMM` package leverages data on vehicle speeds on a set of road links at various points in time to predict the probility distribution of travel time on a given sequence of links, or trip.  The availability of full probability distributions allows better risk management during trip planning.  The package implements a version of the TRIP method proposed in Woodard et al. (2017).

## Example data set: `tripset`

We will use the example data set `tripset` to explore the functionalities offered by `traveltimeHMM`.  Data such as that found in `tripset` can be generated, for instance, using GPS data collected from mobile phones.  In the present case, the data comes from a research project and has therefore been fully anonymised.

```{r}
library(traveltimeHMM)
head(tripset)
```

Travel data is organized around the notions of *trips* and *links*.  Links are road segments each with a well-defined beginning and end points and which can be traversed.  A vehicle performs a trip when it travels from a start point to an end point through a sequence of links.  Thus trips can be considered as ordered sequences of links.  The data set includes data for a collection of trips.

- Field `tripID` contains each trip's ID, whereas field `linkID` contains the IDs of each link making up a trip.  Both fields need to be numerical.  It is assumed that, in the data set, all trips are grouped together and all links of a given trip appear in the order in which they are traversed.  *No verification is performed to that effect*.

- Field `timeBin` refers to the time when the traversal of a given link occurred.  Time bins should reflect as much as possible time periods of the week encompassing similar traffic classes.  In `tripset` we find five time bins: `Weekday`, `MorningRush`, `EveningRush`, `EveningNight` and `Weekendday`.  Field `time` served as a basis for determining the time bin for each link traversal, as illustrated in table 1.

```{r echo = FALSE, results='asis'}
y <- c("Mon - Fri outside rush hour", "Mon - Fri, 7AM - 9AM", "Mon - Fri, 3PM - 6PM", "Sat 9AM - 9PM + Sun 9AM - 7PM", "Sat - Sun otherwise")
x <- c("Weekday", "MorningRush", "EveningRush", "Weekendday", "EveningNight")
z <- data.frame(cbind(y, x))
names(z) <- c("Period of the week", "Time bin")
knitr::kable(z, caption = "Table 1 - Example from `tripset` : Time bin by period of the week")
```

- Field `logspeed` contains the natural logarithm of the speed (in km/h) of traversal of each link.

- Fields `traveltime` and `length` refer respectively to the traversal time (in seconds) and length (in meters) of each link.  Those fields are mostly for reference and are not used directly in the package.  Field `time` also falls in this category; however, data in that field is likely to be useful for providing the starting time of a trip to the prediction function.  We'll have more to say about this later.

## References

Woodard, D., Nogin, G., Koch, P., Racz, D., Goldszmidt, M., Horvitz, E., 2017.  "Predicting travel time reliability using mobile phone GPS data".  *Transportation Research Part C*, 75, 30-44.  <http://dx.doi.org/10.1016/j.trc.2016.10.011>

## TO BE CONTINUED

## Old stock - the rest of the document can be deleted

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
